Research log

### 2023-04-16 analysis/formatDataForTuring.Rmd
model-based analysis with orthogonalized regressors

    matrix <- matrix(c(post_subset,prior_subset,eig_subset),length(mask_indices),3)
    if (Rank(mask_indices)==3) {
    orthogonalized_matrix <- gramSchmidt(matrix)$Q

>>>>

    matrix <- matrix(c(post_subset,prior_subset,eig_subset),length(mask_indices),3)
    if (Rank(matrix)==3) {
    orthogonalized_matrix <- gramSchmidt(matrix)$Q

### 2023-04-16 docs/ExploratoryResults.Rmd
model-based analysis with orthogonalized regressors

>>>>

## Model-based analysis

We considered three factors that contribute to letter selections in Hangman:

1. The prior probability of a letter in the English language $p(X)$.
2. The posterior probability that a letter is present in the target word, given the game state $S$ and the category $p(X|S)$.
3. The expected information gain this letter provides regarding the identity of the target word $H(S)-\sum p(x_v)H(S,x_v)$ (where $x_v$ is one potential outcome of guessing the letter $X$, for example, finding it in the second and fifth positions).

To quantify the contribution of each of these factors to letter selection, we fitted a multinomial logistic regression model to the data of each player. We compared the ensuing coefficients in 'pretend' and 'non-pretend' conditions.

```{r hangman-load-coefficients, echo=FALSE}

#get list of subjects
E4.data_for_turing <- read.csv('../analysis/dataForTuring/E4norm.csv')
E4.subjects <- E4.data_for_turing$subj_id%>%unique()

#initialize
E4.model_coefficients_raw <- data.frame()

#fill the array
for (i_subj in 1:length(E4.subjects)) {
  E4.model_coefficients_raw <- E4.model_coefficients_raw %>%
    rbind(read.csv(sprintf('../analysis/model_fits/E4/subj%d_normalized.csv',i_subj)) %>%
            mutate(subj_id=E4.subjects[i_subj]) %>%
            dplyr::select(subj_id,term,estimate))
}

# Organize in a more convenient form
# non-pretend dataframe

E4.np_coefficients <- E4.model_coefficients_raw %>% 
  group_by(subj_id) %>% 
  summarize(test_part = 'nonpretend',
            prior = estimate[term == 'coef_prior'],
            post = estimate[term == 'coef_post'],
            eig = estimate[term == 'coef_eig'])

E4.p_coefficients <- E4.model_coefficients_raw %>% 
  group_by(subj_id) %>% 
  summarize(test_part='pretend',
            prior = estimate[term == 'coef_prior'] +
              estimate[term == 'coef_prior_delta'],
            post = estimate[term == 'coef_post'] +
              estimate[term == 'coef_post_delta'],
            eig = estimate[term == 'coef_eig'] +
              estimate[term == 'coef_eig_delta'])

E4.model_coefficients <- E4.np_coefficients %>%
  rbind(E4.p_coefficients)
```

```{r hangman-coefficients_plot, echo-FALSE, cache=TRUE, fig.width=10,fig.height=5}

E4.model_coefficients %>%
  gather('term','coefficient',prior,post,eig) %>%
  mutate(test_part = factor(test_part, levels=c('nonpretend','pretend'),
                            labels=c('non-pretend','pretend')),
         term = factor(term, levels = c('prior','post','eig'), labels = c('prior','posterior','expected information gain'))) %>%
  ggplot(aes(x=coefficient,fill=test_part, color=test_part))+
  geom_histogram( alpha=0.6, position = 'identity', bins=100) +
  scale_fill_manual(values=c("#69b3a2", "#404080")) +
  scale_color_manual(values=c("#146464", "#404080")) +
  labs(x='coefficient (a.u.)', y='number of players')+
  theme(legend.position=c(0.10,0.85)) +
  scale_x_continuous(limits=c(-2,5)) +
  facet_wrap(~term, nrow=3)+
  # scale_x_reverse() +
  theme_classic();

ggsave('../docs/figures/E4coefs.png',width=5,height=4.5)
```

All three terms had a positive effect in both conditions. Prior probability (base rate) had a stronger effect in pretend games (`r apa_print(E4.model_coefficients %>% dplyr::select(subj_id,test_part,prior)%>%spread(test_part,prior)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$full_result`). Hit posterior probability showed no such difference, and had a similar effect on players' decisions in both conditions (`r apa_print(E4.model_coefficients %>% dplyr::select(subj_id,test_part,post)%>%spread(test_part,post)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$full_result`). Finally, expected information gain had a weaker effect in pretend games (`r apa_print(E4.model_coefficients %>% dplyr::select(subj_id,test_part,eig)%>%spread(test_part,eig)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$full_result`).

```{r hangman-load-coefficients-orth, echo=FALSE}

#get list of subjects

#initialize
E4.model_coefficients_orth_raw <- data.frame()

#fill the array
for (i_subj in 1:length(E4.subjects)) {
  E4.model_coefficients_orth_raw <- E4.model_coefficients_orth_raw %>%
    rbind(read.csv(sprintf('../analysis/model_fits/E4/subj%d_orth.csv',i_subj)) %>%
            mutate(subj_id=E4.subjects[i_subj]) %>%
            dplyr::select(subj_id,term,estimate))
}

# Organize in a more convenient form
# non-pretend dataframe

E4.np_orth_coefficients <- E4.model_coefficients_orth_raw %>% 
  group_by(subj_id) %>% 
  summarize(test_part = 'nonpretend',
            prior = estimate[term == 'coef_prior'],
            post = estimate[term == 'coef_post'],
            eig = estimate[term == 'coef_eig'])

E4.p_orth_coefficients <- E4.model_coefficients_orth_raw %>% 
  group_by(subj_id) %>% 
  summarize(test_part='pretend',
            prior = estimate[term == 'coef_prior'] +
              estimate[term == 'coef_prior_delta'],
            post = estimate[term == 'coef_post'] +
              estimate[term == 'coef_post_delta'],
            eig = estimate[term == 'coef_eig'] +
              estimate[term == 'coef_eig_delta'])

E4.model_coefficients_orth <- E4.np_orth_coefficients %>%
  rbind(E4.p_orth_coefficients)
```

Given colinearity between model predictors, we suspected that the effects of expected information gain may be driven by posterior probability effects.  To isolate unique contributions of prior probability and expected information gain, we repeated the same analysis, but this time serially orthogonalizing the features (in the order [posterior, prior, expected information gain], and using the Gram Schmidt algorithm) prior to fitting the multinomial regression model. This time, expected information gain had no significant effect on decision in non-pretend (`r  apa_print(E4.model_coefficients_orth%>%filter(test_part=='nonpretend')%>%pull(eig)%>%t.test())$statistic`) and pretend games (`r  apa_print(E4.model_coefficients_orth%>%filter(test_part=='pretend')%>%pull(eig)%>%t.test())$statistic`), rendering it possible that the previously observed effects were confounded with prior and posterior probability. In contrast, both prior and posterior probability had a significant positive effect, with prior probability having a stronger effect in pretend games (`r apa_print(E4.model_coefficients_orth %>% dplyr::select(subj_id,test_part,prior)%>%spread(test_part,prior)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$full_result`).

```{r hangman-coefficients_plot_orth, echo-FALSE, cache=TRUE, fig.width=10,fig.height=5}

E4.model_coefficients_orth %>%
  gather('term','coefficient',prior,post,eig) %>%
  mutate(test_part = factor(test_part, levels=c('nonpretend','pretend'),
                            labels=c('non-pretend','pretend')),
         term = factor(term, levels = c('prior','post','eig'), labels = c('prior','posterior','expected information gain'))) %>%
  ggplot(aes(x=coefficient,fill=test_part, color=test_part))+
  geom_histogram( alpha=0.6, position = 'identity', bins=100) +
  scale_fill_manual(values=c("#69b3a2", "#404080")) +
  scale_color_manual(values=c("#146464", "#404080")) +
  labs(x='coefficient (a.u.)', y='number of players', title='Orthogonalized vectors')+
  theme(legend.position=c(0.10,0.85)) +
  scale_x_continuous(limits=c(-2,5)) +
  facet_wrap(~term, nrow=3)+
  # scale_x_reverse() +
  theme_classic();

ggsave('../docs/figures/E4coefs_orth.png',width=5,height=4.5)
```
